---
title: "Birthdays: Probability, Statistics and Data Visualisation"
author: "Zak Varty"
date: "Spring 2019"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true 
    logo: lancaster_black1.png
    runtime: shiny 
---  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Questions:

 - What is the probability that two or more people in this room share a birthday?


 - What assumptions are you making to arrive at your answer? 
```{r, fig.align='center'}
#countdown::countdown_timer(60)
```

## Typical Assumptions

- No twins, triplets, etc.  (birth dates are independent)
- All days are equally probable. 
- There are 365 days in the year. (Not always true)  
- $$ \begin{align} \text{Pr}(n \text{  distinct birthdays}) &= \frac{365}{365} \times \frac{364}{365} \times \dots \times \frac{365 - (n -1)}{365} \\ 
& \\
&= \prod_{i=0}^{n-1} \left\{\frac{365-i}{365}\right\}. \end{align}$$
- Pr(Shared birthday) = 1 - Pr($n$ distinct birthdays).  

## Plotting this Solution
<div class="centered">
```{r}
n<- 40
N <- seq(1:n)
pr <- rep(NA,n)
for(i in N){
  pr[i] <- prod((365: (365-(i-1)))/365)
}
plot(x = N,y = pr,type = 'h', ylab = 'Pr( n distinct birthdays )')
```
</div>

## Plotting this Solution
```{r}
library(shiny)

# Define UI for app that draws the pmf ----
ui <- fluidPage(

  # App title ----
  titlePanel(""),

  # Sidebar layout with input and output definitions ----
  sidebarLayout(

    # Sidebar panel for inputs ----
    sidebarPanel(

      # Input: Slider for the number of bins ----
      sliderInput(inputId = "students",
                  label = "Number of students:",
                  min = 1,
                  max = 70,
                  value = 30)

    ),

    # Main panel for displaying outputs ----
    mainPanel(

      # Output: Histogram ----
      plotOutput(outputId = "probPlot")

    )
  )
)
```

```{r}
# Define server logic required to draw a histogram ----
server <- function(input, output) {

  # pmf of sharing birthdays ----
  # with requested number of students highlighted and probability as main
  # This expression that generates the pmf is wrapped in a call
  # to renderPlot to indicate that:
  #
  # 1. It is "reactive" and therefore should be automatically
  #    re-executed when inputs (input$students) change
  # 2. Its output type is a plot
  output$probPlot <- renderPlot({

    x    <- 1:100 
    pr <- rep(NA,100)
    for(i in 1:100){
      pr[i] <- prod((365: (365-(i-1)))/365)
    }
    col <- rep(1,100)
    col[input$students] <- 2
    
    plot(x = x,
         y = 1- pr,
         type = 'h',
         ylab = 'Pr( shared birthday )',
         col = col,
         main = paste(round(1- pr[input$students], digits = 8)),
         lwd = 4)
  }
 )
}
```

```{r}
shinyApp(
  ui = ui,
  server = server,
  options = list(height = 500)
)
```


## Are all days really equally likely?

```{r load birth data}
#library('readxl')
birthsbydate <- readxl::read_excel("birthsbydate1995to2016.xls", 
    sheet = "Data", na = "z", skip = 4, n_max = 366)
```


```{r split Date into day and month}
library('tidyr')
birthsbydate <- birthsbydate %>%
  separate(Date, into = c("day", "month"), sep = "/")
```

```{r}
birthsbyday <- birthsbydate %>%
  dplyr::mutate(total = rowSums(.[,-(1:2)],na.rm = TRUE)) %>%
  dplyr::select(day,month,total) #%>%
  # dplyr::mutate(day = as.integer(day)) %>%
  # dplyr::mutate(month = as.integer(month)) $>$
  # dplyr::mutate(total = as.integer(total))

births_mat <- spread(data = birthsbyday,value = total,key = day)
```

```{r, fig.align="center"}

birthsbyday <- birthsbyday %>% 
  dplyr::mutate(sds = (total- mean(total,na.rm = TRUE))/sd(total,na.rm = TRUE)) %>%
  dplyr::mutate(day = factor(day)) %>%
  dplyr::mutate(month = factor(month)) %>% 
  dplyr::mutate(month = factor(x = month, levels = rev(levels(month))))
  

ggplot2::ggplot(data = birthsbyday, 
                ggplot2::aes(x= day, 
                             y= month,
                             fill= total)) + 
  ggplot2::geom_tile() + 
  ggplot2::ggtitle("UK births by day of year: Total Count 1995-2016") +
  ggplot2::scale_fill_gradient(low="black", high="red") 
```

## Correcting for Feb 29
```{r correcting for Feb29, fig.align="center"}
occurances <- rep(22,366)
occurances[31+29] <- 6 #number of leap years (1996,2000,...,2016).
birthsbyday <- birthsbyday %>%
  dplyr::mutate(occurs =  occurances) %>%
  dplyr::mutate(rel_count = total/occurs)


ggplot2::ggplot(data = birthsbyday, 
                ggplot2::aes(x= day, 
                             y= month,
                             fill= rel_count)) + 
  ggplot2::geom_tile() + 
  ggplot2::ggtitle("UK births by day of year: Adjusted Mean Count 1995-2016") +
  ggplot2::scale_fill_gradient(low="black", high="red") + 
  ggplot2::labs(fill = "Mean \n Count")
```

## A different view of the same data
```{r A different view, fig.align="center"}

birthsbydate <- dplyr::mutate(.data = birthsbydate, doy = 1:366)
birthsbyday <- dplyr::mutate(.data = birthsbyday, doy = 1:366)

ggplot2::ggplot(data = birthsbyday) +
  ggplot2::geom_line(mapping = ggplot2::aes(x = doy,y = rel_count)) + 
  ggplot2::ylab("mean birth count") + 
  ggplot2::xlab("day of year")
```


## Handling holidays 

- Only emergency treatments on Christmas, Boxing day and New Years Day. 
- Treat these days separately. (Skip over hypothesis tests here) 

- $$\widehat p = \frac{\text{relative count}}{\text{total relative count}}$$ $$ \widehat{p}_{01/01} = 0.00237 \approx \frac{1}{421}$$
$$ \widehat{p}_{25/12} = 0.00214 \approx \frac{1}{467}$$
$$ \widehat{p}_{26/12} = 0.00204 \approx \frac{1}{490}$$
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=TRUE}
nyd <- which(birthsbyday$month == '01'& birthsbyday$day =='01')
boxday <- which(birthsbyday$month == '12'& birthsbyday$day =='26')
xmas <- which(birthsbyday$month == '12'& birthsbyday$day =='25')
leap <- which(birthsbyday$month == '02'& birthsbyday$day =='29')
phat <- function(dayofyear){
  birthsbyday$rel_count[dayofyear] / sum(birthsbyday$rel_count)
} 
phat(c( xmas, boxday, nyd))
1/phat(c( xmas, boxday, nyd))
```

## What about the non-holidays?
```{r warning=FALSE, fig.align="center"}
non_hols_byday <- birthsbyday
nyd <- which(birthsbyday$month == '01'& birthsbyday$day =='01')
boxday <- which(birthsbyday$month == '12'& birthsbyday$day =='26')
xmas <- which(birthsbyday$month == '12'& birthsbyday$day =='25')
leap <- which(birthsbyday$month == '02'& birthsbyday$day =='29')
hols <- c(nyd,boxday,xmas, leap)
non_hols_byday[hols,c(3,4,6)]<- NA

ggplot2::ggplot(data=non_hols_byday) + 
  ggplot2::geom_line(mapping = ggplot2::aes(x = doy, y = rel_count)) +
  ggplot2::xlab("day of year") + 
  ggplot2::ylab("mean birth count")
```

## But we have all of this information 
```{r message=FALSE, warning=FALSE, fig.align="center"}
non_hols <- birthsbydate 
nyd <- which(non_hols$month == '01'& non_hols$day =='01')
boxday <- which(non_hols$month == '12'& non_hols$day =='26')
xmas <- which(non_hols$month == '12'& non_hols$day =='25')
leap <- which(non_hols$month == '02'& non_hols$day =='29')
hols <- c(nyd,boxday,xmas, leap)
non_hols[hols,3:25] <- NA

tidynon_hols <- gather(data = non_hols, key = year,value = births, `2016`:`1995`)
ggplot2::ggplot(data = tidynon_hols) +
  ggplot2::geom_point(mapping = ggplot2::aes(x = doy,y = births, col = year)) + 
  ggplot2::ylab("birth count") + 
  ggplot2::xlab("day of year")
```


## What are we testing?


$$ H_0: \text{all non-holidays have the same mean birth count.}$$

$$vs.$$

$$H_1: \text{At least one day has a different mean.}$$

## Bootstrap confidence interval - idea
- Under $H_0$ and the observationd being i.i.d. the observed counts could have happened on any day
- We can build another day that we have not observed based on our data
- Build lots of these days and calculate the mean of each one
- Use the quantiles of these e.g. $(\bar x_{0.025}, \bar x_{0.975})$ to construct a confidence interval.



## Bootstrap confidence interval - procedure 
- Sample 22 non holiday birth counts at random, with replacement.
- Calculate their mean.
- Add this to a vector of means. 
- Repeat until we have 10,000 means. 
- Calculate the desired quantiles of the mean vector. 
- Compare to observations. 

## Bootstrap confidence interval - code
```{r echo=TRUE}
boot_data <- as.vector(as.matrix(non_hols[,3:24]))
boot_sample <- rep(NA,22)

boot_size <- 1e5
boot_means  <- rep(NA,boot_size) 

for(i in 1:boot_size){
  boot_sample <- sample(boot_data, size = 22, replace = TRUE)
  boot_means[i] <- mean(boot_sample,na.rm = TRUE)
}

boot_CI <- quantile(boot_means, probs = c(0.025,0.975))
boot_CI
```

## Bootstrap confidence interval - results

```{r warning=FALSE, fig.align="center"}
non_hols_byday <- birthsbyday
nyd <- which(birthsbyday$month == '01'& birthsbyday$day =='01')
boxday <- which(birthsbyday$month == '12'& birthsbyday$day =='26')
xmas <- which(birthsbyday$month == '12'& birthsbyday$day =='25')
leap <- which(birthsbyday$month == '02'& birthsbyday$day =='29')
hols <- c(nyd,boxday,xmas, leap)
non_hols_byday[hols,c(3,4,6)]<- NA

ggplot2::ggplot(data=non_hols_byday) + 
  ggplot2::geom_line(mapping = ggplot2::aes(x = doy, y = rel_count)) +
  ggplot2::xlab("day of year") + 
  ggplot2::ylab("mean birth count") + 
  ggplot2::geom_hline(yintercept = boot_CI, linetype = 'dashed', color = 'red')
```

## Other questions

- Is there dependence on the previous count(s)?
- Is there an effect from the day of week?
- Is there an trend over time?
- If you want to learn more: https://pudding.cool/2018/04/birthday-paradox/


## Takeaways
- Statistics is about understanding the assumptions. 
- The more assumptions you make the more you can conclude.
- Display choices impact what is seen and what is hidden. 
- A statistician's work is never done. 


## Thank you 
Any quick questions?

(I'll be around at coffee for longer ones!)
